Process-based phenotypes: an application to the Cellular Phenotype Ontology

### 2011-11-18 trunk/paper/drinking.tex

In English language, we may say that an object has a {\em frequency}
or {\em rate} as if these were attributes of objects at time
points. For example, that my heart {\em now} beats with a frequency of
80 beats per minute, or that a car is moving with a speed of 280
kilometres per minute {\em at a particular point in time} (e.g., as
observed with a speed camera) may be considered attributes of the
objects, not attributes of processes. In our analysis, rates as
attributes of objects can be explicitly defined using rates of
processes. For example, the heart beating frequency of a particular
heart $h$ at a time point $t$ is the frequency of heart beating
processes in which $h$ participates at an infinitesimally small
process around $t$ and within a reference process. The reference
process is necessary in order to obtain a value for a frequency or
rate even if the process in question is not occurring. For example, a
rate of cardiac muscle contraction can be determined even when no
cardiac muscle contraction is occurring in virtue of a reference
process with a rate of cardiac muscle contraction.

@@

\subsection{Cell Phenotype Ontology: implementation and evaluation}
We were faced with two choices for implementing the CPO: we could
either implement a pre-composed ontology in which all classes and
their definitions are pre-generated according to the patterns we
define, or we could develop an annotation software that enables the
selection of our process phenotype patterns based on the current
structure of the GO.  To maximize the utility and compatibility of the
CPO, and to provide stable identifiers for its concepts, we selected
the first strategy and developed a software to automatically generate
a pre-composed ontology from the GO.

@@

In the CPO, identifiers are of the form {\tt CPO:XXGGGGGGG}, where
{\tt GGGGGGG} is the seven-digit identifier of the GO class on which
the CPO class is based, and {\tt XX} is a prefix that identifies the
type of pattern we applied to the GO class. For example, based on the
class {\em Apoptotis} ({\tt GO:0006915}), we generate the CPO classes
{\em Abnormality of Apoptosis}, {\em Abnormality of single occurrence
  of apoptosis} and {\em Abnormality of regulation of apoptosis}.
We use the prefixes {\tt 12}, {\tt 14} and {\tt 15} for each of the
corresponding phenotype patterns, and consequently generate the class
identifiers {\tt CPO:120006915}, {\tt CPO:140006915} and {\tt
  CPO:150006915}.

>>>>

We may say that an object has a {\em frequency} or {\em rate} as if
these were attributes of objects at time points. For example, that my
heart {\em now} beats with a frequency of 80 beats per minute, or that
a car is moving with a speed of 280 kilometres per minute {\em at a
  particular point in time} (e.g., as observed with a speed camera)
may be considered attributes of the objects, not attributes of
processes. In our analysis, rates as attributes of objects can be
explicitly defined using rates of processes. For example, the heart
beating frequency of a particular heart $h$ at a time point $t$ is the
frequency of heart beating processes in which $h$ participates at an
infinitesimally small process around $t$ and within a reference
process. The reference process is necessary in order to obtain a value
for a frequency or rate even if the process in question is not
occurring. For example, a rate of cardiac muscle contraction can be
determined even when no cardiac muscle contraction is occurring in
virtue of a reference process with a rate of cardiac muscle
contraction.

@@

\subsection{Cell Phenotype Ontology: implementation}
We were faced with two choices for implementing the CPO: we could
either implement a pre-composed ontology in which all classes and
their definitions are pre-generated according to the patterns we
define, or we could develop an annotation software that enables the
selection of our process phenotype patterns based on the current
structure of the GO.  To maximize the utility and compatibility of the
CPO, and to provide stable identifiers for its concepts, we selected
the first strategy and developed a software to automatically generate
a pre-composed ontology from the GO.

@@

In the CPO, identifiers contain two components and are of the form
{\tt CPO:XXGGGGGGG}, where {\tt GGGGGGG} is the seven-digit identifier
of the GO class on which the CPO class is based, and {\tt XX} is a
prefix that identifies the type of phenotype pattern that is applied
to the GO class. For example, based on the class {\em Apoptosis} ({\tt
  GO:0006915}), we generate the CPO classes {\em Abnormality of
  Apoptosis}, {\em Abnormality of single occurrence of apoptosis} and
{\em Abnormality of regulation of apoptosis}.  We use the prefixes
{\tt 12}, {\tt 14} and {\tt 15} for each of the corresponding
phenotype patterns, and consequently generate the class identifiers
{\tt CPO:120006915}, {\tt CPO:140006915} and {\tt CPO:150006915}. As
long as the GO maintains its identifier for the {\em Apoptosis} class,
the identifiers in the CPO will remain stable even when it is
regenerated.

We use the labels of GO classes to automatically generate class labels
for phenotype classes as well as textual definitions for classes in
the CPO. For example, the label of the class for increased number of
occurrences of {\em Apoptosis} is {\em Increased frequency of
  occurrences of Apoptosis}, and its textual definition states:
``An increased frequency of occurrences of Apoptosis is a phenotype of
entities in which the number of occurrences of Apoptosis within a
given time period is increased in comparison to a reference
process.''
